Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan

### 2024-08-19 analysis/paper/paper.qmd
work on the material section

<!--archaeological materials-->
The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005]. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials. Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through a regional exchange network, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan has continuously participated in a regional trade network closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@]. Moreover, the use of foreign goods appears to have some cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].

<!--archaeological remains-->
There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area. The animals include deer, pigs, cows, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but limited to a household scale, according to historical records [@]. Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources. Also, those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities. According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources.

# Mateirals and methods

>>>>

<!--archaeological materials-->
The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005]. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials. Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through exchange, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan has continuously participated in regional trade networks closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@Wang2007]. Moreover, the use of foreign goods appears to have cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].

<!--archaeological remains-->
There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area. The animals include deer, pigs, cows, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but limited to a household scale, according to historical records [@]. Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources. Those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities. According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources. The resource use during the periods of external impacts still need further examination to better understands the changes over time.

# Mateirals and methods
To explore changes in the subsistence strategies before and after the European arrival with associated long-distance trade network, I examined zooarchaeological and botanical evidence related to economic activities in a colonial context from Kiwulan. The faunal and botanical assemblages analyzed in this study are from 40 units located in the residential area excavated during the first season (2001-2006) at Kiwulan. The 40 excavation units were selected due to their relativity intact chronological contexts and spatially continuous area. Because deerskin was the main commodity in local regional exchange since the 14th century and further largely demanded during the Dutch colonization period, deer bones and their butchery patterns are especially focused.
